Make any stupid plot you want

### 2021-02-18 draft/sections/intro.tex
reworking intro + math with Z&K glyphs

\section{Background}
Visual algorithms that display information are designed such that structure of data is assumed according to Tory and Möller  \cite{toryRethinkingVisualizationHighLevel2004}. Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}. This idea can be seen in modern general purpose visualization architecture,  where many libraries are either defined around specific data structures or make implicit assumptions about the data or the types of visualizations they will support. In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

@@

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create. It is also the wrong abstraction layer since the grammer exposes specific ways to compose parts together to build a chart, but not ways to build new parts\cite{wongsuphasawatNavigatingWideWorld2021}.

@@

There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview}. These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic. This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing. We are proposing a functional framework instead in large part to clearly encapsulate these separate responsibilities of a visualization tool. In turn, this should make it easier to reuse components in the building of new tools.

\note{Add somehow: Software Design Patterns for Information Visualization (heer, agrawala)}

@@

We propose that the Du Bois in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express. This is because while the Du Bois figures are not the common scatter, bar, or line plot \cite{friendlyBriefHistoryData2008}, they conform to the constraint that a visualization is a mapping from data to visuals. In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph. The visual representations in the Du Bois figures are in proportion to the quantitative data being represented, so the a chart is faithful according to Tufte's Lie Principal\cite{tufteVisualDisplayQuantitative2001}. The properties of the representation match the properties of the information being represented, so they are fairly understandable according to Norman's Naturalness Principal\cite{norman_things_smart}. As with the Du Bois data portraits, it is fundemental that any architecture tool we build ensures that the data properties match the visual properties.

@@

The visual properties were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who also described the types of data that paired with the visual properties shown in figure~\ref{fig:intro_retinal_variables}. It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}. These visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond. Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}. As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools. By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.

\subsection{Contribution}
\note{should I mention the categorical framework here too?}
This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes

>>>>

\section{Background}
Visual algorithms that display information are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}. Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}. This assumption is evident in the information visualization software design patterns categorized by Heer and Agrawala\cite{HeerSoftware2006} and can often be found in modern general purpose visualization libraries.  In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

@@

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.  Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements.

@@

There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic. This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing. We are proposing a functional framework instead in large part to clearly encapsulate these separate responsibilities of a visualization tool. In turn, this should make it easier to reuse components in the building of new tools.

@@

We propose that the Du Bois visualizations in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express. This is because while the Du Bois figures are not the common scatter, bar, or line plot \cite{friendlyBriefHistoryData2008}, they conform to the constraint that a visualization is a mapping from data to visuals. In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph. The visual representations in the Du Bois figures are in proportion to the quantitative data being represented, so the a chart is faithful according to Tufte's Lie Principal\cite{tufteVisualDisplayQuantitative2001}. The properties of the representation match the properties of the information being represented, so they are fairly understandable according to Norman's Naturalness Principal\cite{norman_things_smart}. As with the Du Bois data portraits, it is fundamental that any architecture tool we build ensures that the data properties match the visual properties.

@@

The visual properties were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who also described the types of data that paired with the visual properties shown in figure~\ref{fig:intro_retinal_variables}. It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}. Vickers et al. propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.

The visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables}. Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}. As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools. By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.

\subsection{Contribution}
This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes

### 2021-02-18 draft/sections/math.tex
reworking intro + math with Z&K glyphs

We can describe a mark \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} as \vmark\ with input of all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
\begin{equation}
\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
\end{equation}
where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1]. We define the mark as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$

\begin{equation}
    \begin{tikzcd}
        \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
        \end{tikzcd}
    \label{eq:mark}
\end{equation}
such that for every mark there is at least one corresponding section on \dbase. We define a glyph as a composite of multiple \vmark\, in keeping with Munzner's definition of the glyph as an object with internal structure arising from multiple marks\cite{munznerVisualizationAnalysisDesign2014}.

@@

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
    \caption{The torus \dtotal is unraveled into a simplacial complex of 2 faces \dbase. Transition functions are defined on the edges of \dbase such that surface can be glued back into the torus.
    \note{add cross sections a and b to ring and color same as edges in complex}}
    \label{fig:triangle_torus}
\end{figure}
An advantage of fiber bundles is the ability to encode data continuity that is more complex than points, a line or a plane. For example, the tori in figure~\ref{fig:triangle_torus} can be unraveled into a simplacial complex of two triangles with labeled edges. Transition functions $\delta$ are defined on the edges such that $a$ can be glued to $a^\prime$ and $b$ to $b^\prime$ to reconstruct the torus. This simplacial complex is then used as the base space encoding the continuity of data that lies in the torus. A constraint on the transition functions is that the monoid actions on the fibers on the edges of \dtotal\ are commutative $\monoid*\dfiber\mapsto \delta(\monoid\dfiber)= \monoid*\delta(\dfiber)$

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
    \caption{Many non-trivial spaces can be made locally trivial by dividing \dtotal into locally trivial subspaces and defining transition functions between the edges on \dbase for how to glue the two subspaces such that the \dsection are continuous.}
    \label{fig:data_base_transition}
\end{figure}
Another advantages of triangulization is that it provides a way to encode non-trivial structures such as the mobius strip\cite{MobiusStripNLab}. As shown in figure ~\ref{fig:data_base_transition}, one way of making the mobius strip trivial is to seperate it into two spaces $\dtotal_1$ and $\dtotal_2$ and then define transition functions that specify that the edges of $\dtotal_1$ need to be reversed to line up with $\dtotal_2$ such that the sections along the edges meet. As with the torus, the transition functions must preserve monoid commutativity. While we generally do not deal with complex or non-trivial bundles, the ability to implement them native to the model allows for a great deal of expressivity in encoding data. This is especially important when developing a tool that needs to be general enough to support any number of very domain specific types of data.

>>>>

 We can formally describe a glyph as \vmark\, with the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
\begin{equation}
\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
\end{equation}
where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1]. We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
\begin{equation}
    \begin{tikzcd}
        \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
        \end{tikzcd}
    \label{eq:mark}
\end{equation}
such that for every glyph there is at least one corresponding section on \dbase. This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}. The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.

@@

In general simplacial complexes, in these prototypes we implement graphs.
